Text-mining clinically relevant cancer biomarkers for curation into the CIViC database

### 2021-06-25 paper/paper.Rmd
Tweaks to paper and paper code

**Methods**: To aid in this curation and provide the greatest coverage for these databases, particularly CIViC, we propose the use of text mining approaches to extract these clinically relevant biomarkers from all available published literature. To this end, a group of cancer genomics experts annotated biomarkers and their clinical associations discussed in 800 sentences and achieved good inter-annotator agreement. We then used a supervised learning approach to construct the CIViCmine knowledgebase (http://bionlp.bcgsc.ca/civicmine/)

@@

**Conclusions**: Through integration with CIVIC, we provide a prioritised list of curatable clinically relevant cancer biomarkers as well as a resource that is valuable to other knowledgebases and precision cancer analysts in general. All data is publically available and distributed with a Creative Commons Zero license.

@@

## Introduction

@@

The main objective of this project was to identify frequently discussed cancer biomarkers that fit the CIViC evidence model but are not yet included in the CIViC knowledgebase. We developed an information extraction-based method to extract key parts of the evidence item: cancer type, gene, drug (where applicable) and the specific evidence type from published literature. This allows us to count the number of mentions of specific evidence items in abstracts and full-text articles and compare against the CIViC knowledgebase. This paper will present our methods to develop this resource, known as CIViCmine (http://bionlp.bcgsc.ca/civicmine/). The main contributions of this work are an approach for knowledgebase construction that could be applied to many areas of biology and medicine, a machine learning method for extracting complicated relationships between four entity types, and extraction of relationships across the largest possible publically-accessible set of abstracts and full-text articles. This resource, containing `r paper.biomarkerCount` gene-cancer associations with clinical relevance, is valuable to all cancer knowledgebases to aid their curation and also as a tool for precision cancer analysts searching for evidence supporting biomarkers not yet included in any other resource.

@@

@@

To manage this compute and provide infrastructure for easy updating with new publications in all three corpora, we made use of the updated PubRunner infrastructure (paper in preparation - https://github.com/jakelever/pubrunner). This allows for easy distribution of the work across a compute cluster. The resulting data was then pushed to Zenodo for perpetual and public hosting (https://doi.org/10.5281/zenodo.1156241). The data is released with a Creative Commons Public Domain (CC0) license so that other groups can easily make use of it.

@@

biomarkerComparisonPlot.withTitle <- arrangeGrob(biomarkerComparisonPlot,top='(a)')
pmidComparisonPlot.withTitle <- arrangeGrob(pmidComparisonPlot,top='(b)')
cgiPredictivePlot.withTitle <- arrangeGrob(cgiPredictivePlot,top='(c)')
cgiPredisposingPlot.withTitle <- arrangeGrob(cgiPredisposingPlot,top='(d)')
oncoKBPredictivePlot.withTitle <- arrangeGrob(oncoKBPredictivePlot,top='(e)')

comparisonPlot <- arrangeGrob(biomarkerComparisonPlot,
            pmidComparisonPlot,
            cgiPredictivePlot,
            cgiPredisposingPlot,
            oncoKBPredictivePlot,
            ncol=2)

@@

The CIViCmine app, accessible at http://bionlp.bcgsc.ca/civicmine, and freely accessible associated data provide a valuable addition to the precision oncology informatics community. CIViCmine can be used to assist curation of other precision cancer knowledgebases and can be used directly by precision cancer analysts to search for biomarkers of interest. As this resource will be updated monthly with the latest research, it will constantly change as new cancer types and drug names enter the lexicon. We anticipate that the methods described can be used in other biomedical domains and that the resources provided will be valuable to the biomedical text mining and precision oncology fields.

@@

The complete datasets are available in the Zenodo repository (https://doi.org/10.5281/zenodo.1472826) with the September 2019 release used for this paper (https://doi.org/10.5281/zenodo.3441694). The data can also be viewed through the web viewer (http://bionlp.bcgsc.ca/civicmine/) and subsets of the data can be downloaded there. The code for the analysis and web viewer is available at GitHub (https://github.com/jakelever/civicmine/).

@@

## Author Contributions

>>>>

**Methods**: To aid in this curation and provide the greatest coverage for these databases, particularly CIViC, we propose the use of text mining approaches to extract these clinically relevant biomarkers from all available published literature. To this end, a group of cancer genomics experts annotated sentences that discussed biomarkers with their clinical associations and achieved good inter-annotator agreement. We then used a supervised learning approach to construct the CIViCmine knowledgebase.

@@

**Conclusions**: Through integration with CIVIC, we provide a prioritised list of curatable clinically relevant cancer biomarkers as well as a resource that is valuable to other knowledgebases and precision cancer analysts in general. All data is publically available and distributed with a Creative Commons Zero license. The CIViCmine knowledgebase is available at http://bionlp.bcgsc.ca/civicmine/.

@@

@@

@@

The main objective of this project was to identify frequently discussed cancer biomarkers that fit the CIViC evidence model but are not yet included in the CIViC knowledgebase. We developed an information extraction-based method to extract key parts of the evidence item: cancer type, gene, drug (where applicable) and the specific evidence type from published literature. This allows us to count the number of mentions of specific evidence items in abstracts and full-text articles and compare against the CIViC knowledgebase. We present our methods to develop this resource, known as CIViCmine (http://bionlp.bcgsc.ca/civicmine/). The main contributions of this work are an approach for knowledgebase construction that could be applied to many areas of biology and medicine, a machine learning method for extracting complicated relationships between four entity types, and extraction of relationships across the largest possible publically-accessible set of abstracts and full-text articles. This resource, containing `r paper.biomarkerCount` gene-cancer associations with clinical relevance, is valuable to all cancer knowledgebases to aid their curation and also as a tool for precision cancer analysts searching for evidence supporting biomarkers not yet included in any other resource.

@@

To manage this compute and provide infrastructure for easy updating with new publications in all three corpora, we made use of the updated PubRunner infrastructure (paper in preparation - https://github.com/jakelever/pubrunner). This allows for easy distribution of the work across a compute cluster. The resulting data was then pushed to Zenodo for perpetual and public hosting [@jake_lever_2019_3441694]. The data is released with a Creative Commons Public Domain (CC0) license so that other groups can easily make use of it.

@@

biomarkerComparisonPlot.withTitle <- arrangeGrob(biomarkerComparisonPlot,bottom='Evidence Items')
pmidComparisonPlot.withTitle <- arrangeGrob(pmidComparisonPlot,bottom='PubMed IDs')
cgiPredictivePlot.withTitle <- arrangeGrob(cgiPredictivePlot,bottom='Predictive')
cgiPredisposingPlot.withTitle <- arrangeGrob(cgiPredisposingPlot,bottom='Predisposing')
oncoKBPredictivePlot.withTitle <- arrangeGrob(oncoKBPredictivePlot,bottom='Predictive')

comparisonPlot <- arrangeGrob(biomarkerComparisonPlot.withTitle,
            pmidComparisonPlot.withTitle,
            cgiPredictivePlot.withTitle,
            cgiPredisposingPlot.withTitle,
            oncoKBPredictivePlot.withTitle,
            ncol=2)

@@

## Conclusions

The CIViCmine resource, accessible at http://bionlp.bcgsc.ca/civicmine, and freely available associated data provide a valuable addition to the precision oncology informatics community. CIViCmine can be used to assist curation of other precision cancer knowledgebases and can be used directly by precision cancer analysts to search for biomarkers of interest. As this resource will be updated monthly with the latest research, it will constantly change as new cancer types and drug names enter the lexicon. We anticipate that the methods described can be used in other biomedical domains and that the resources provided will be valuable to the biomedical text mining and precision oncology fields.

@@

The complete datasets are available in the Zenodo repository (https://doi.org/10.5281/zenodo.1472826) with the September 2019 release used for this paper [@jake_lever_2019_3441694]. The data can also be viewed through the web viewer (http://bionlp.bcgsc.ca/civicmine/) and subsets of the data can be downloaded there. The code for the analysis and web viewer is available at GitHub (https://github.com/jakelever/civicmine/).

@@

## Authors' Contributions
